\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-04-20 honors_thesis/ppsn_paper/paper.tex
Down to 11 pages on the PPSN paper, but still some work to go

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{chu-2006-intelligent-parallel-pso, jin-2005-pso-antenna-designs,
parsopoulos-2004-parallel-vector-evaluated-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{mostaghim-2006-multi-objective-pso-on-grids,
venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes. Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once, then
presents a comparison of the speculative algorithm with the original PSO.  In
\sec{conclusion} and \sec{future} we conclude and discuss future work.

@@

The value of the objective function at $\ppos_{t+1}$ is given by:
\begin{align}
\label{eq:val2update}
	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
	\ \ofunc(\specpos_{t+1}^{c})
\end{align}

@@

\subsection{Topology}
\label{sec:topology}

The number of speculative evaluations needed per particle depends on the number
of neighbors each particle has.  The number of update cases in a topology where
each particle has $n$ neighbors is $2(n+1)$; there are two possibilities for
updates to $\pbest$ (updated by the particle itself and not updated), and $n+1$
possibilities for updates to $\nbest$ (updated by each neighboring particle and
not updated).  When the particle is also a neighbor to itself, as is always the
case in commonly used topologies, one of the cases can be eliminated, as a
particle cannot be the source of an update to its neighborhood best while not
also updating its personal best.  Thus we have $2(n+1)-1$, or $2n+1$,
speculative evaluations per particle.  In a swarm with $p$ particles and $n$
neighbors per particle, $(2n+1)p$ speculative evaluations are needed.

Because the number of speculative evaluations depends on the number of
neighbors a particle has, the choice of topology is an important one.  The use
of the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  In such cases, we use the Random topology 
which has been shown to
effectively approximate the rapid information flow of the Complete topology.
In our implementation of the random topology there are to randomly chosen neighbors
(chosen anew at every iteration)~\cite{mcnabb-2009-large-particle-swarms}.

In the rest of this
section we move from the mathematics of our method to its implementation.
First we discuss the relatively easy case of a centralized parallel PSO
algorithm with a master computer and many slaves.  In such an architecture, the
master keeps track of all necessary information with only trivial message
passing needed.  Then we discuss the more complicated case of a distributed
algorithm, where each particle is on its own and needs to send and receive
messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

\subsection{Terminology}

@@

We separate each iteration of PSO into several steps.  First there is the
motion step, where a particle updates its position and velocity.  Then a
particle's position is evaluated, and the particle updates its current value
and its personal best.  Finally, a particle gets information from its neighbors
and updates its neighborhood best.

@@

\subsection{Implementation}

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-2004-parallel-models-for-pso}.  To perform
speculative evaluation in such an architecture, the master generates the
positions to evaluate speculatively as in \eqref{eq:defpcasegen}.  After having
the slaves evaluate the objective function at all necessary positions, the
master then decides which position to accept for each particle, as in
\eqref{eq:val2updatelong}.  The outline of the speculative evaluation in PSO (SEPSO) algorithm is given in
\alg{centralized}.

@@

In a distributed parallel PSO implementation, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for those
particles.  Messages such as
values and positions for the neighborhood best are sent between processors.
There may still be some machine that collects information from all of the
particles and outputs the result of the algorithm, though that machine's
importance is much less than in centralized algorithms.
In some distributed frameworks rounds of message passing can be expensive.
A N\:aive implementation of speculative evaluation
uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
perform speculative evaluation in PSO with only one round of communication per
two iterations. 
Dynamic topologies like the random topology discussed above, further complicate the
needed message passing.
However, the message passing implementation is not particularly
interesting, but is tedious thus, in the interest of space, is omitted.

@@

As discussed in \sec{topology}
where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as complete leads to an explosion of speculative
evaluations.
For the standard PSO baseline we have included experiments with the
ring and the random topologies, both with two neighbors, as well as for the complete topology.
It is important to note however, that in many cases the complete topology is not a
practical alternative in the context of large
swarms on large cluster where where the messaging complexity is $O(n^2)$ and can
overwhelm the system.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere1}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round. Note that PSO Complete requires $O(n^2)$ messaging
  and may not be practical in many cases.}
  \label{fig:sphere1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank2}
\end{figure}

@@

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome and how
many processors are available.  It is likely that the best thing to do with
those functions is to use all processors to add particles to the swarm and do
no speculative evaluation.

\section{Future Work}
\label{sec:future}

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.  For
example, our technique required speculatively evaluating every possible next
position of each particle.  The number of speculative evaluations could
potentially be greatly reduced by using statistics for which branch of motion
particles most often follow.  Only the subset of the most probable next
locations would then be evaluated.  Using an asynchronous PSO variant may also
be helpful in this case, allowing particles to be at different iterations in
case the correct branch was not guessed and needs to be evaluated separately.

Another possible advancement would be to reduce communication costs by dropping
some of the inter-processor messages in distributed frameworks.  The PSO
algorithm would have to be adapted to work with less information, perhaps not
requiring that each processor have completely accurate information about what
neighboring particles are doing.

Still another possibility is to view the set of possible evaluations as an
infinite tree.  Each level in the tree is an iteration, and the branches on
each level represent the possible next positions of the particle.  One need not
simply evaluate all of one level for each particle---perhaps going two or three
iterations ahead on the most likely path would perform better.

We have mentioned that this technique has been previously used to parallelize
simulated annealing.  Any other optimization algorithm that only depends on
current sampling positions when computing the next position to sample can be
parallelized with this technique.  In particular, genetic algorithms produce
future generations by combining individuals from the current generation.  With
a large population size there would be an unwieldy amount of possible future
individuals, but the potential exists to modify the algorithm to use some kind
of speculative evaluation.  A related but quite distinct technique for reducing
computation in evolutionary algorithms was proposed
in~\cite{poli-2006-backward-chaining-evolutionary-algorithms}.

Large parallel clusters are often required to successfully optimize practical
modern problems.  To properly use PSO with such clusters, a balance needs to be
made between using processors to increase the swarm size and using them to
increase the speed of the algorithm, or for perhaps as yet unconsidered
techniques.  This work is a first step in that direction.

>>>>

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{chu-2006-intelligent-parallel-pso,
schutte-2004-parallel-global-optimization-with-pso} and asynchronous algorithms
\cite{venter-2005-parallel-pso-asynchronous-evaluations}.  Parallelizing the
evaluation of the objective function can also be done in some cases, though
that is not an adaption of the PSO algorithm itself and thus is not the focus
of this paper.

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes. Error bars show median and
  10th and 90th percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once, then
presents a comparison of the speculative algorithm with the original PSO.  In
\sec{conclusion} we conclude.

@@

@@

\subsection{Implementation}
\label{sec:implementation}

The number of speculative evaluations needed per particle depends on the number
of neighbors each particle has.  In a swarm with $p$ particles and $n$
neighbors per particle, $(2n+1)p$ speculative evaluations are necessary (each
additional neighbor adds two rows to Table~\ref{tab:evals}).  This dependence
on the number of neighbors necessitates a wise choice of topology.  The use of
the Complete topology, where every particle is a neighbor to every other
particle, would require $O(p^2)$ speculative evaluations per iteration.  It is
much more desirable to have a sparse topology, where $O(np)$ is much smaller
than $O(p^2)$.  However, some functions are better optimized with the Complete
topology and the quick spread of information it entails than with sparse
topologies.  In such cases, we use the Random topology described in \sec{pso}.

@@

@@

The outline of the speculative evaluation in PSO (SEPSO) algorithm is given in
\alg{centralized}.

@@

@@

As discussed in \sec{implementation},
where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion of speculative
evaluations.
For the standard PSO baseline we have included experiments with the
ring and the random topologies, both with two neighbors, as well as for the complete topology.
It is important to note however, that in many cases the complete topology is not a
practical alternative in the context of large
swarms on large cluster where where the messaging complexity is $O(n^2)$ and can
overwhelm the system.

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{sphere1}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round. Note that PSO Complete requires $O(n^2)$ messaging
  and may not be practical in many cases.}
  \label{fig:sphere1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank2}
\end{figure}

@@

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-2009-large-particle-swarms}.  For those functions the correct
use of processors cannot be easily prescribed and depends largely on the
desired outcome and how many processors are available.  It is likely that the
best thing to do with those functions is to use all processors to add particles
to the swarm and do no speculative evaluation.
